### 2018-02-14 submission/manuscript.Rmd
update fig calls in discussion

The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, tumor formation. The results that we have presented show that both the genera- and OTU-based select and full models, for indviduals with carcinoma, had similar AUCs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma turmors. Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. Additionally, how the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 6]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F]. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F]. The majority of the colon tissue-based results seem to be study specific with many of the top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could explain why these results seem to be more sporadic than the stool results.

@@

The associations between the microbiota and indivdials with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas. Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation. In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 7].

>>>>

The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, tumor formation. The results that we have presented show that both the genera- and OTU-based select and full models, for indviduals with carcinoma, had similar AUCs [Figure 2 & 3]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma turmors. Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. Additionally, how the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 4]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 4]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S4E & F]. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S4B-C & E-F]. The majority of the colon tissue-based results seem to be study specific with many of the top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could explain why these results seem to be more sporadic than the stool results.

@@

The associations between the microbiota and indivdials with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas. Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation. In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].

### 2018-02-14 submission/manuscript.tex
update fig calls in discussion

The data presented herein support the importance of select genera for
carcinoma, but not necessarily adenoma, tumor formation. The results
that we have presented show that both the genera- and OTU-based select
and full models, for indviduals with carcinoma, had similar AUCs
{[}Figure 4 \& 5{]}. This suggests that an interplay between a select
number of potentially protective and exacerbating microbes within the GI
community could be crucial for carcinoma formation. Importantly, it
suggests that there may be key members of the GI community that should
be studied further to potentially help reduce the risk of carcinoma
tumor formation. Conversely, using the present data, it is clear that
new approaches may be needed to identify members of the community
associated with adenoma turmors. Regardless of sample type and whether a
full or select model was used, our Random Forest models consistently
performed poorly. Yet, the step-wise decrease in diversity suggests that
the adenoma-associated community is not normal but has changed subtly
{[}Figure 1{]}. This change in diversity, at this early stage of
disease, could be focal to the adenoma itself. Additionally, how the
host interacts with these subtle changes at early stages of the disease
could be what leads to a thoroughly dysfunctional community that is
supportive of tumorgenesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across studies {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be in the top 10 most
important variables across the different studies for both individuals
with adenoma and carcinoma {[}Figure 6{]}. These data suggest that
whether the non-resident bacterium is \emph{Fusobacteria} or
\emph{Peptostreptococcus} may not be as important as how these bacteria
interact with the changing resident community. Based on these
observations, it is possible to hypothesize that small changes in
community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding
of how the microbiota may be associated with tumors. Generally, the full
OTU-based models of unmatched and matched colon tissue samples were
concordant with stool samples showing that GI resident microbes were the
most prevalent in the top 10 most important variables across study
{[}Figure S7E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
only mouth-associated bacteria consistently present in the top 10 most
important variables of the full carcinoma stage models {[}Figure S7B-C
\& E-F{]}. The majority of the colon tissue-based results seem to be
study specific with many of the top 10 taxa being present only in a
single study. Additionally, the presence of genera associated with
contamination, within the top 10 most important variables for the genera
and OTU models is worrying. The low bacterial biomass of tissue samples
coupled with potential contamination could explain why these results
seem to be more sporadic than the stool results.

@@

The associations between the microbiota and indivdials with adenoma
tumors are inconclusive, in part, because many studies may not be
powered effectively to observe small effect sizes. None of the studies
analyzed were properly powered to detect a 10\% or lower change between
cases and controls. The results within our meta-analysis suggest that a
small effect size may well be the scope in which differences
consistently occur between controls and those with adenomas. Future
studies investigating adenoma tumors and the microbiota need to take
power into consideration to reproducibly study whether the microbiota
contributes to polyp formation. In contrast to adenoma stage of disease,
our observations suggest that most studies analyzed have sufficient
power to detect many changes in the carcinoma-associated microbiota
because of large effect size differences between cases and controls
{[}Figure 7{]}.

>>>>

The data presented herein support the importance of select genera for
carcinoma, but not necessarily adenoma, tumor formation. The results
that we have presented show that both the genera- and OTU-based select
and full models, for indviduals with carcinoma, had similar AUCs
{[}Figure 2 \& 3{]}. This suggests that an interplay between a select
number of potentially protective and exacerbating microbes within the GI
community could be crucial for carcinoma formation. Importantly, it
suggests that there may be key members of the GI community that should
be studied further to potentially help reduce the risk of carcinoma
tumor formation. Conversely, using the present data, it is clear that
new approaches may be needed to identify members of the community
associated with adenoma turmors. Regardless of sample type and whether a
full or select model was used, our Random Forest models consistently
performed poorly. Yet, the step-wise decrease in diversity suggests that
the adenoma-associated community is not normal but has changed subtly.
This change in diversity, at this early stage of disease, could be focal
to the adenoma itself. Additionally, how the host interacts with these
subtle changes at early stages of the disease could be what leads to a
thoroughly dysfunctional community that is supportive of tumorgenesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across studies {[}Figure 4{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be in the top 10 most
important variables across the different studies for both individuals
with adenoma and carcinoma {[}Figure 4{]}. These data suggest that
whether the non-resident bacterium is \emph{Fusobacteria} or
\emph{Peptostreptococcus} may not be as important as how these bacteria
interact with the changing resident community. Based on these
observations, it is possible to hypothesize that small changes in
community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding
of how the microbiota may be associated with tumors. Generally, the full
OTU-based models of unmatched and matched colon tissue samples were
concordant with stool samples showing that GI resident microbes were the
most prevalent in the top 10 most important variables across study
{[}Figure S4E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
only mouth-associated bacteria consistently present in the top 10 most
important variables of the full carcinoma stage models {[}Figure S4B-C
\& E-F{]}. The majority of the colon tissue-based results seem to be
study specific with many of the top 10 taxa being present only in a
single study. Additionally, the presence of genera associated with
contamination, within the top 10 most important variables for the genera
and OTU models is worrying. The low bacterial biomass of tissue samples
coupled with potential contamination could explain why these results
seem to be more sporadic than the stool results.

@@

The associations between the microbiota and indivdials with adenoma
tumors are inconclusive, in part, because many studies may not be
powered effectively to observe small effect sizes. None of the studies
analyzed were properly powered to detect a 10\% or lower change between
cases and controls. The results within our meta-analysis suggest that a
small effect size may well be the scope in which differences
consistently occur between controls and those with adenomas. Future
studies investigating adenoma tumors and the microbiota need to take
power into consideration to reproducibly study whether the microbiota
contributes to polyp formation. In contrast to adenoma stage of disease,
our observations suggest that most studies analyzed have sufficient
power to detect many changes in the carcinoma-associated microbiota
because of large effect size differences between cases and controls
{[}Figure 5{]}.
